Appendix for 'Framing Effects in the Wild

### 2022-09-29 paper/Manuscript_Framing.Rmd
keep tex, final proofread

---
title: "Limits of Media Effects"
subtitle: "How Politicisation Shapes the Pace of Attitudinal Change"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    keep_tex: true
    toc: false
    fig_caption: yes
    number_sections: false
    includes:
      in_header: preamble.tex
# subtitle: Manuscript prepared for the 12th Annual Conference of the European Political Science Association in Prague, Czech Republic
fontsize: 12pt
geometry: margin=1.5in
abstract: "What are the limits of the news media's ability to shape public opinion? I argue that politicization conditions the responsiveness of individual attitudes to changes in the information environment. Exploiting a rare change in the migration framing of the biggest German tabloid *Bild*, I estimate the precise impact of this shift on immigration attitudes using large-scale panel data. Despite a 42 percent increase of the emphasis of crime in migration coverage, I find a robust null effect on immigration attitudes and a number of related variables. Further analyses suggest that media outlets' ability to manipulate public opinion in the short term seems to be limited to brief opportunity windows during the politicisation of a given issue."
keywords: "Public Opinion, Media Effects, News Framing, Opinion Formation, Emphasis Framing, Conditions of Media Effects."
bibliography: C:/Users/nicol/Dropbox/Studium/BibTex/PhD.bib
nocite: |
  @CologneReport2016
indent: true
---

@@

In the remainder of this paper, I will argue that *Bild* missed a critical window of opportunity to shape public opinion. As readers' immigration attitudes had crystallized following the politicization of immigration in 2015, their immigration attitudes remained resistant to a change in the tabloid's migration framing. I support my argument with evidence presented in three steps: first, I show that *Bild* indeed increasingly emphasized crime-related news in its migration coverage, following an exogenously timed change of the editor-in-chief. I then exploit this natural experiment to provide evidence that this change did little to affect readers' immigration attitudes or related variables. In a last step, I trace different explanations for the absence of this effect, suggesting that the pace of attitudinal change severely decreased following the politicization of the issue in 2015.

To assess whether the migration framing in *Bild* really changed, I develop an approach to measure framing in political texts. I apply this framework to a dataset of 2.5 million articles from the major German daily newspapers. Using supervised classifiers, I first select migration-related coverage, to then estimate the prevalence of crime content within this coverage. This analysis suggests that following the change of the editor-in-chief, the emphasis of crime in migration content increased by over 40 percent.

To estimate the effect of this changing framing on immigration attitudes, I employ panel data from the German Longitudinal Election Study, fielded in the period of interest [@GLES2017Panel]. I use a *difference-in-differences* (DiD) design to estimate the impact of the increased emphasis of crime in the coverage of migration and refugees. Results show that the changing framing did not result in more restrictive attitudes towards immigration, nor did it affect the association of crime with migration, views on integration, or the perceived importance of the immigration issue.

Lastly, I explore different explanations why this shift in migration framing did not bring about attitudinal changes. First, I assess whether readers with initially liberal migration attitudes selected out of the newspaper in response to the changing portrayal of migrants. Reported readership of respondents with liberal migration attitudes remains at similar levels as that of readers with restrictive migration attitudes. Second, I investigate whether readers with many different sources of information are less reactive to the changing coverage and mask the effect. I find no effect heterogeneity across the reported consumption of newspaper, TV and online news, ruling out the possibility that readers with more diverse information hide an effect among exclusive *Bild* readers. Third, I test whether attitudes change after longer-term exposure. I find small effects after more than one and a half years, suggesting a very slow pace of attiudinal change. Lastly, I assess whether attitudes crystallized after the politicization of immigration in 2015. Using data from a long-term online panel [@GLES2018LangfristPanel], I show that the average individual change in immigration attitudes was substantially lower in the period 2016-2018 compared to the period 2014-2016.

@@

Lastly, the evidence speaks to the study of framing effects. Fifteen years ago, Donald Kinder called for observational research on framing effects: "Experimental results can always be questioned on their generalizability, and framing effects are no exception. [...] A more balanced reading of frame effects requires methodological diversification, experiments and studies oriented to the world outside" [-@Kinder2007, 157]. By precisely identifying the nature of the change in news coverage, this study answered this call. The null effects found suggest that framing effects do not easily generalize from the experimental context.

The paper proceeds as follows: I provide a brief review of the expectancy-value model of attitude formation, which informs my hypotheses and content analysis, before discussing potential limits of framing effects in real-world environments. I then present the research design, content analysis, and a discussion of expected and minimal effect size of interest, before turning to the estimation of the effect of changing immigration framing on immigration attitudes and trace different explanations for the observed resistance. In the conclusion, I outline the implications of this model for framing, media, and public opinion research, as well as the limitations of this study.

@@

"Framing is the process by which a communication source, such as a news organisation, defines and constructs a political issue or public controversy" [@Nelson1997Rally]. Different forms of frames have been conceptualized. When I talk about 'frames' here, I refer to *emphasis frames*. These frames emphasize certain topics in relation to an issue, guiding the recipient to think about the issue with those considerations in mind that are promoted by the frame [@Leeper2020, 152]. Many other forms of framing exist, such as equivalence frames [@Kahneman1979prospect], episodic frames [@Iyengar2005], or generic frames [@DeVreese2005]. The focus on emphasis frames in this study is owed to the nature of the case, as well as the generalized theory regarding the effect of emphasis frames on attitudes about any issue (as opposed to generic and episodic frames) and their common use in news content (which is not the case for equivalence frames).

@@

Research on framing has shown for decades that citizens' attitudes and their issue associations are highly reactive to minor changes in the emphasis of specific considerations [@Chong2007b; @Chong2010; @Druckman2012Bias; @Kinder1990; @Nelson1996; @Nelson1997Rally; @sniderman2004structure; for reviews see @Busby2019; @Druckman2001; @Leeper2020]. However, it remains unclear how well these experimental findings translate into the real world. Below, I outline three key assumptions in the generalization of experimental estimates.

First, taking experimental effects as estimates of the real-world impact of framing means making strong assumptions about citizens' exposure to frames: while framing studies usually simply present news articles or other forms of political content to respondents, news content in the real world competes with other news and entertainment content to reach its audience. Changes in news framing itself might affect how likely readers are to consume news content: readers might be less likely to consume frames challenging their existing attitudes and reject news content containing such frames [and vice versa seek out attitudinally congruent information, @Arceneaux2010; @Druckman2012Bias]. This process of 'selective exposure/avoidance' [@Bennett2008], or a 'confirmation bias' [@Taber2006] brings with it a major endogeneity problem making purely correlational evidence rather uninformative. As most framing studies do not offer choice of content to respondents, engagement remains an assumption when taking experimental studies as evidence of framing effects in the real world [for an exception see @Druckman2012Bias].

@@

In the summer of 2015, Germany became sanctuary for around 800,000 refugees who had fled war, violence, and famine in Syria and elsewhere. The German public debated how to deal with the newly arriving and for the first time since the post-war years, a radical right party was likely to win representation in the national parliament. Surprisingly, given its traditional populist style and right-wing takes on migration, the major tabloid newspaper *Bild* initially maintained a somewhat balanced coverage of refugees, which – although still portraying migrants more negatively than other outlets [@Maurer2019] – even started a campaign called "#RefugeesWelcome" to gather support for the newly arriving\footnote{See \url{https://web.archive.org/web/20160114154402/https://www.bild.de/news/topics/fluechtlingshilfe/wir-helfen-buehne-42385428.bild.html}.}. This was often connected to chief editor Kai Diekmann, who hosted a refugee family in his home [@BR2015Diekmann].

@@

To measure migration framing in the news, I collected 2.5 million newspaper articles from the websites of the most important German broadsheets *Frankfurter Allgemeine Zeitung (FAZ), Spiegel Online (SPON), Süddeutsche Zeitung (SZ), Die Tageszeitung (TAZ), Die Welt*, as well as *Bild* for the period 2013-2019. These outlets reach a substantial share of the German public, with *Bild* distributing over 1.7 million print copies and registering over 550 million web impressions at the time (see section 2.1 in the online appendix for more information). Following the value-expectancy model, I identify crime frames in migration coverage in a two-step process: first, I use supervised machine learning to identify content about migration. Second, I identify content about crime *within* migration coverage by training and applying another supervised classifier. This approach closely resembles the value-expectancy model outlined [above](#framing-effects), as I first identify coverage about the *issue*, before identifying the consideration emphasized in the discussion of this issue.

@@

ATT <- single_model$coefficients[["postTRUE:treatTRUE"]]
t_ATT <- single_model$coeftable["postTRUE:treatTRUE", "t value"]
p_ATT <- single_model$coeftable["postTRUE:treatTRUE", "Pr(>|t|)"]

@@

# Conclusion: The Ebb and Flow of Attitudinal Change

What are the limits of media power? While the answer to this question has profound implications about the extent of elite influence in contemporary democracies, we currently lack evidence about the scope conditions of media effects. This study addresses this gap by presenting evidence from a rare shift in the migration coverage of Germany's largest tabloid, *Bild*. The presented results are robust evidence that readers' immigration attitudes remained surprisingly unaffected. Estimates of attitudinal change are indistinguishable from zero and remain below minimal effect sizes in a number of specifications, dependent variables, and subgroups.

@@

While this evidence might seem like a contradiction to the idea that attitudes change *most* during the politicization of an issue, it is useful to take a longitudinal perspective. The model I propose views political issues as defined in public discourse. Once an issue emerges in public debate (usually because of an external event), problem definitions, causal explanations, and potential solutions are collectively debated. Political actors struggle to get their frames considered. Only a small subset of these frames will get to dominate public debate, shaping citizens' understanding of the issue. This collective definition will be resistant to attempts of framing and thus stabilize attitudes. Attitudinal change hence accelerates during the politicization of an issue, as citizens update their perceptions. Once a stable collective issue definition emerges, individuals become resistant to further information about the issue.

This model also speaks to a large body of media effects research. It suggests that the power of the news media to shape public opinion is more limited than previously thought. Even when elites try, they might not always be able to manipulate their audience's political attitudes.  While studies on media effects has shown that public opinion *can* be affected by changing news exposure [@DellaVigna2007; @Foos2022; @Grossman2022; @Levendusky2022], this study took a first step to investigate under which conditions this process takes place.

This study has several limitations, which should be addressed by future research. The argument in this paper applies to liberal democracies in general, but I present evidence from Germany, which is a Western European democracy with a relatively high-quality press, large public broadcasters, and a moderately involved electorate. While the experimental evidence from the US, Switzerland, and Denmark discussed above supports the idea that this might be a more general phenomenon, future observational studies should assess longitudinal panel data in other contexts. Similarly, migration might be a 'special' issue, meaning that citizens form different attitudes differently on this issue [@Kustov2021]. Lastly, my analysis of the 'crystallization' of attitudes is limited by the available data. To observe the ac- and deceleration of attitudinal change relative to issue's politicization, future research could exploit panel data from different contexts and on different issues, but ultimately more longer-term panel data tracking issue attitudes will be necessary. This is especially challenging, as attitudinal measures are usually added to survey questionnaires only *after* the politicization of these issues. However, to get a full picture of the process of attitude formation, it is necessary to track the same issues irrespective of their current relevance in day-to-day politics. Otherwise, researchers will only observe issue attitudes after the relevant processes of attitude formation have taken place.

>>>>

---
title: "Limits of Media Effects"
subtitle: "How Politicisation Shapes the Pace of Attitudinal Change"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    keep_tex: true
    toc: false
    fig_caption: yes
    number_sections: false
    includes:
      in_header: preamble.tex
fontsize: 12pt
geometry: margin=1.5in
abstract: "What are the limits of the news media's ability to shape public opinion? I argue that politicization conditions the responsiveness of individual attitudes to changes in the information environment. Exploiting a rare change in the migration framing of the biggest German tabloid *Bild*, I estimate the precise impact of this shift on immigration attitudes using large-scale panel data. Despite a 42 percent increase of the emphasis of crime in migration coverage, I find a robust null effect on immigration attitudes and a number of related variables. Further analyses suggest that media outlets' ability to manipulate public opinion seems to be limited to brief opportunity windows during the politicisation of a given issue."
keywords: "Public Opinion, Media Effects, News Framing, Opinion Formation, Emphasis Framing, Conditions of Media Effects."
bibliography: C:/Users/nicol/Dropbox/Studium/BibTex/PhD.bib
nocite: |
  @CologneReport2016
indent: true
---

@@

I argue that *Bild* missed a critical window of opportunity to shape public opinion. As readers' immigration attitudes had crystallized following the politicization of immigration in 2015, their immigration attitudes remained resistant to a change in the tabloid's migration framing. I support my argument with evidence presented in three steps: first, I show that *Bild* increasingly emphasized crime-related news in its migration coverage, following an exogenously timed change of the editor-in-chief. I then exploit this quasi-experiment to provide evidence that this change did little to affect readers' immigration attitudes or related variables. In a last step, I trace different explanations for the absence of this effect, suggesting that the pace of attitudinal change severely decreased following the politicization of the issue in 2015.

To assess whether the migration framing in *Bild* really changed, I develop an approach to measure framing in political texts. I apply this framework to a dataset of 2.5 million articles from the major German daily newspapers. Using supervised classifiers, I first identify migration-related coverage, to then estimate the prevalence of crime content within this coverage. This analysis suggests that following the change of the editor-in-chief, the emphasis of crime in migration content increased by over 40 percent.

To estimate the effect of this changing framing on immigration attitudes, I employ panel data from the German Longitudinal Election Study, fielded in the period of interest [@GLES2017Panel]. I use a *difference-in-differences* (DiD) design to estimate the impact of the increased emphasis of crime in the coverage of migration and refugees. Results show that the changing framing did not result in more restrictive attitudes towards immigration, nor did it affect the association of crime with immigration, views on integration, the perceived importance of the immigration issue, or support for the radical right.

Lastly, I explore different explanations why this shift in migration framing did not bring about attitudinal changes. First, I assess whether readers with initially liberal migration attitudes selected out of the newspaper in response to the changing portrayal of migrants. Reported readership of respondents with liberal migration attitudes remains at similar levels as that of readers with restrictive migration attitudes. Second, I investigate whether readers with many different sources of information are less reactive to the changing coverage and mask the effect. I find no effect heterogeneity across the reported consumption of newspaper, TV and online news, ruling out the possibility that readers with more diverse information hide an effect among exclusive *Bild* readers. Third, I test whether attitudes change after longer-term exposure. I find small effects after more than one and a half years, suggesting a very slow pace of attitudinal change. Lastly, I assess whether attitudes crystallized after the politicization of immigration in 2015. Using data from a long-term online panel [@GLES2018LangfristPanel], I show that the average individual change in immigration attitudes was substantially lower in the period 2016-2018 compared to the period 2014-2016.

@@

Lastly, the evidence speaks to the study of framing effects. Fifteen years ago, Donald Kinder called for observational research on framing effects: "Experimental results can always be questioned on their generalizability, and framing effects are no exception. [...] A more balanced reading of frame effects requires methodological diversification, experiments and studies oriented to the world outside" [-@Kinder2007, 157]. By precisely identifying framing in news coverage, as well as its effects on public opinion, this study answered his call. The null effects found suggest that framing effects do not easily generalize from the experimental context.

The paper proceeds as follows: I provide a brief review of the expectancy-value model of attitude formation, which informs my hypotheses and content analysis, before discussing potential limits of framing effects in real-world environments. I then present the research design, content analysis, and a discussion of expected and minimal effect size of interest, before turning to the estimation of the effect of interest. The last empirical part traces different explanations for the observed resistance. In the conclusion, I outline the implications of my model, the limitations of the study, as well as avenues for future research.

@@

"Framing is the process by which a communication source, such as a news organisation, defines and constructs a political issue or public controversy" [@Nelson1997Rally]. Different forms of frames have been conceptualized. When I talk about 'frames' here, I refer to *emphasis frames*. These frames emphasize certain topics in relation to an issue, guiding the recipient to think about the issue with those considerations in mind that are promoted by the frame [@Leeper2020, 152]. Many other forms of framing exist, such as equivalence frames [@Kahneman1979prospect], episodic frames [@Iyengar2005], or generic frames [@DeVreese2005]. The focus on emphasis frames in this study is owed to the generalized theory regarding the effect of emphasis frames on attitudes about any issue (as opposed to generic and episodic frames) and their common use in news content (which is not the case for equivalence frames).

@@

Research on framing has shown for decades that citizens' attitudes and their issue associations are highly reactive to minor changes in the emphasis of specific considerations [@Chong2007b; @Chong2010; @Druckman2012Bias; @Kinder1990; @Nelson1996; @Nelson1997Rally; @sniderman2004structure; for reviews see @Busby2019; @Druckman2001; @Leeper2020]. However, it remains unclear how well these experimental findings translate into citizens' everyday realities. Below, I outline three key assumptions in the generalization of experimental estimates.

First, taking experimental effects as estimates of the real-world impact of framing means making strong assumptions about citizens' exposure to frames: while framing studies usually simply present news articles or other forms of political content to respondents, news content in the real world competes with other news and entertainment content to reach its audience. Changes in news framing itself might affect how likely citizens are to consume news content: readers might be less likely to consume frames challenging their existing attitudes and reject news content containing such frames [and vice versa seek out attitudinally congruent information, @Arceneaux2010; @Druckman2012Bias]. This process of 'selective exposure/avoidance' [@Bennett2008], or a 'confirmation bias' [@Taber2006] brings with it a major endogeneity problem making purely correlational evidence rather uninformative. As most framing studies do not offer choice of content to respondents, engagement remains an assumption when taking experimental studies as evidence of framing effects in the real world [for an exception see @Druckman2012Bias].

@@

In the summer of 2015, Germany became sanctuary for around 800,000 refugees who had fled war, violence, and famine in Syria and elsewhere. The German public debated how to deal with the newly arriving and for the first time since the post-war years, a radical right party was likely to win seats in the national parliament. Surprisingly, given its traditional populist style and right-wing takes on migration, the major tabloid newspaper *Bild* initially maintained a somewhat balanced coverage of refugees, which – although still portraying migrants more negatively than other outlets [@Maurer2019] – even started a campaign called "#RefugeesWelcome" to gather support for the newly arriving\footnote{See \url{https://web.archive.org/web/20160114154402/https://www.bild.de/news/topics/fluechtlingshilfe/wir-helfen-buehne-42385428.bild.html}.}. This was often connected to chief editor Kai Diekmann, who hosted a refugee family in his home [@BR2015Diekmann].

@@

To measure migration framing in the news, I collected 2.5 million newspaper articles from the websites of the most important German broadsheets *Frankfurter Allgemeine Zeitung (FAZ), Spiegel Online (SPON), Süddeutsche Zeitung (SZ), Die Tageszeitung (TAZ), Die Welt*, as well as *Bild* for the period 2013-2019. These outlets reach a substantial share of the German public, with *Bild* distributing over 1.7 million print copies and registering over 550 million web impressions at the time (see section 2.1 in the online appendix for more information). Following the value-expectancy model, I identify crime frames in migration coverage in a two-step process: first, I identify content about migration. Second, I identify content about crime *within* migration coverage. This approach closely resembles the value-expectancy model outlined [above](#framing-effects), as I first identify coverage about the *issue*, before identifying the *consideration* emphasized in the discussion of this issue.

@@

ATT <- single_model$coefficients[["postTRUE:treatTRUE"]]
t_ATT <- single_model$coeftable["postTRUE:treatTRUE", "t value"]
p_ATT <- single_model$coeftable["postTRUE:treatTRUE", "Pr(>|t|))"]

@@

# Conclusion

What are the limits of media power? While the answer to this question has profound implications for our understanding of elite influence in contemporary democracies, we currently lack evidence about the scope conditions of media effects. This study addresses this gap by presenting evidence from a rare shift in the migration coverage of Germany's largest tabloid, *Bild*. The presented results are robust evidence that readers' immigration attitudes remained surprisingly unaffected. Estimates of attitudinal change are indistinguishable from zero and remain below minimal effect sizes in a number of specifications, dependent variables, and subgroups.

@@

While this evidence might seem like a contradiction to the idea that attitudes change *most* during the politicization of an issue, it is useful to take a longitudinal perspective. The model I propose views political issues as defined in public discourse. Once an issue emerges in public debate (usually because of an external event), problem definitions, causal explanations, and potential solutions are collectively debated. Political actors struggle to get their frames considered. Only a small subset of these frames will get to dominate public debate, shaping citizens' understanding of the issue. This collective definition will be resistant to attempts of framing and thus stabilize attitudes. Attitudinal change accelerates during the politicization of an issue, as citizens update their perceptions. Once a stable collective issue definition emerges, individuals become resistant to further information about the issue.

This model also speaks to a large body of media effects research. It suggests that the power of the news media to shape public opinion is more limited than previously thought. Even when elites try, they might not always be able to manipulate their audience's political attitudes.  While studies on media effects have shown that public opinion *can* be affected by changing news exposure [@DellaVigna2007; @Foos2022; @Grossman2022; @Levendusky2022], this study took a first step to investigate under which conditions this process takes place.

This study has several limitations, which should be addressed by future research. The argument in this paper applies to liberal democracies in general, but I present evidence from Germany, which is a Western European democracy with a relatively high-quality press, large public broadcasters, and a moderately involved electorate. While the experimental evidence from the US, Switzerland, and Denmark discussed above supports the idea that this might be a more general phenomenon, future observational studies should assess longitudinal panel data in other contexts. Similarly, migration might be a 'special' issue, meaning that citizens form attitudes differently on this issue [@Kustov2021]. Lastly, my analysis of the 'crystallization' of attitudes is limited by the available data. To observe the ac- and deceleration of attitudinal change relative to issue's politicization, future research should exploit panel data from different contexts and on different issues, but ultimately more longer-term panel data tracking issue attitudes will be necessary. This is especially challenging, as attitudinal measures are usually added to survey questionnaires only *after* the politicization of these issues. However, to get a full picture of the process of attitude formation, it is necessary to track the same issues irrespective of their current relevance in day-to-day politics. Otherwise, researchers will only observe issue attitudes after the relevant processes of attitude formation have taken place.

### 2022-09-29 paper/appendix.Rmd
keep tex, final proofread

---
title: "Appendix for 'Framing Effects in the Wild'"
date: "15 5 2022"
output: 
  bookdown::pdf_document2:
    number_sections: TRUE
toc: TRUE
bibliography: C:/Users/nicol/Dropbox/Studium/BibTex/PhD.bib
---

>>>>

---
title: "Appendix for 'Framing Effects in the Wild'"
date: "15 5 2022"
output: 
  bookdown::pdf_document2:
    number_sections: TRUE
    keep_tex: true
toc: TRUE
bibliography: C:/Users/nicol/Dropbox/Studium/BibTex/PhD.bib
---
